Categories of Optics

### 2018-06-11 paper.tex
Get commented out bits under control
Long and tedious conversion from commas to mids

@@

If $l : S \to M \otimes A$ and $r : M \otimes A' \to S'$, we write $\langle l \mid r \rangle : (S, S') \hto (A, A')$ for the image of this pair in $\Optic_\C((S, S'), (A, A'))$, and say that $M$ is the complement for the representative. Optics will always be written with a crossed arrow $\hto$, to distinguish them from morphisms of $\C$.

\begin{example}
  For any three objects $M, A, A' \in \C$, there is the \emph{tautological} optic \[t_{M,A} : (M \otimes A, M \otimes A') \hto (A, A')\] which is given by $\langle \id_{M \otimes A}\mid \id_{M \otimes A'} \rangle$.
\end{example}

\begin{remark}
  Elements of $\Optic_\C((S, S'), (A, A'))$ have an appealing
  interpretation as string diagrams with a ``hole'' missing. We draw the
  pair $\langle l \mid r \rangle$ as
  \begin{center}
    \input{diagrams/generic-optic.tikz}
  \end{center}
  reading left to right, so the left half of the diagram represents $l$ and the right half $r$. The relation imposed by the coend equates the following two diagrams:
  \begin{center}
    \input{diagrams/coend-relation-left.tikz}
    \qquad
    \input{diagrams/coend-relation-right.tikz}
  \end{center}
  We will therefore omit the vertical cut point between $l$ and $r$ in subsequent diagrams; any choice yields a representative of same optic.
\end{remark}

Two optics can be composed as follows: \todo{I give possibly too much detail here.} We wish to construct a map
\begin{align*}
  &\left(\int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S')\right) \times \left(\int^{M \in \C} \C(R, M \otimes S) \times \C(M \otimes S', R')\right) \\ &
                                                                                                                                                                       \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
The product in $\Set$ preserves colimits, so in particular coends. Using this fact, and the Fubini theorem for coends~\cite[Remark 1.9]{CoendCofriend}, the domain is isomorphic to
\begin{align*}
  \int^{(M, N) \in \C \times \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R').
\end{align*}
So by the universal property of coends, it suffices to construct natural maps
\begin{align*}
  & \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') \\ &
                                                                                                              \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
for each $M, N \in \C$. For this we use the composites
\begin{align*}
  &\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R')\\
  \to \,& \C(N \otimes S, N \otimes M \otimes A) \times \C(N \otimes M \otimes A', N \otimes S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') && \text{(functoriality of $N \otimes  -$)} \\
  \to \,& \C(R, N \otimes  M \otimes A) \times \C(N \otimes M \otimes A', R') && \text{(composition in $\C$)} \\
  \to \,&\int^{M \in \M} \C(R, M \otimes A) \times \C(M \otimes A', R') && \text{($\copr_{N \otimes M}$)}
\end{align*}
In other words, let $\langle l', r' \rangle : (R, R') \hto (S, S')$ and $\langle l, r \rangle : (S, S') \hto (A, A')$ with $M$ the complement for $\langle l', r' \rangle$. The composite is then: $\langle l, r \rangle \circ \langle l', r' \rangle = \langle (M l)l', r'(Mr)\rangle$.

@@

The identity optic $(S, S') \hto (S, S')$ is given by $\langle \lambda^{-1}_S, \lambda_{S'} \rangle$, where $\lambda_S : I \otimes S \to S$ is the left unitor for $S$:
\begin{center}
  \input{diagrams/identity-optic-full.tikz}
\end{center}
This dashed line above the diagram represents the unit object. As is standard, we will elide the unit object and unitors in future diagrams unless they are necessary. The diagram for the identity becomes:
\begin{center}
  \input{diagrams/identity-optic.tikz}
\end{center}

@@

  We check associativity of composition equationally. As in the definition of composition, the Fubini theorem for coends allows us to choose representatives for all three optics simultaneously. Suppose we have representatives of three optics $\langle l_1, r_1\rangle : (R, R) \hto (S, S')$ and $\langle l_2, r_2 \rangle : (S, S') \hto (A, A')$ and $\langle l_3, r_3 \rangle : (A, A') \hto (B, B')$, that have complements $M$, $N$ and $P$ respectively. Then:
  \begin{align*}
    (\langle l_3, r_3 \rangle \circ \langle l_2, r_2 \rangle) \circ \langle l_1, r_1 \rangle
    &= \langle (N \otimes l_3)l_2, r_2(N \otimes r_3) \rangle \circ \langle l_1, r_1 \rangle \\
    &= \langle (M \otimes ((N \otimes l_3)l_2))l_1, r_1(M \otimes (r_2(N \otimes r_3))) \rangle \\
    &= \langle (M \otimes N \otimes l_3)(M \otimes l_2)l_1, r_1(M \otimes r_2)(M \otimes N \otimes r_3) \rangle \\
    &= \langle l_3, r_3 \rangle \circ (\langle (M \otimes l_2)l_1, r_1(M \otimes r_2) \rangle) \\
    &= \langle l_3, r_3 \rangle \circ (\langle l_2, r_2 \rangle \circ \langle l_1, r_1 \rangle)
  \end{align*}

  For the unit laws, suppose we have $\langle l, r \rangle : (S, S') \hto (A, A')$ with representative $M$. We calculate:
  \begin{align*}
    \id_{A, A'} \circ \langle l, r\rangle
    &= \langle \lambda^{-1}_A, \lambda_{A'} \rangle \circ \langle l, r\rangle \\
    &= \langle (M \otimes \lambda^{-1}_A) l, r (M\otimes  \lambda_{A'})\rangle \\
    &= \langle (\lambda^{-1}_M \otimes  A) l, r (\lambda_M \otimes A')\rangle \\
    &= \langle l, r (\lambda_M \otimes A') (\lambda^{-1}_M \otimes A')\rangle \\
    &= \langle l, r \rangle  \\
    \langle l, r \rangle \circ \id_{S, S'}
    &= \langle l, r \rangle \circ \langle \lambda^{-1}_S, \lambda_{S'}\rangle  \\
    &= \langle (I \otimes l)\lambda^{-1}_S, \lambda_{S'} (I \otimes r) \rangle \\
    &= \langle (\lambda^{-1}_M \otimes S)l, r (\lambda_{M} \otimes S') \rangle \\
    &= \langle l, r (\lambda_{M} \otimes S')(\lambda^{-1}_M \otimes S') \rangle \\
    &= \langle l, r \rangle
  \end{align*}
  Where in both cases we have used the naturality of $\lambda$ and coend relations to move $\lambda^{-1}_M$ to the right-hand side of the coend, where it may be cancelled.
\end{proof}

@@

\begin{proposition}
  There is a functor $\iota : \C \times \C^\op \to \Optic_\C$, which is given on objects by $\iota(S, S') = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \langle \lambda_A^{-1} f, g \lambda_{A'} \rangle$.
\end{proposition}
\begin{proof}
  Graphically, this is:
  \begin{center}
    \input{diagrams/iota.tikz}
  \end{center}

  This preserves identities, as the identity on an object $(S, S')$ in $\Optic_\M$ is defined to be exactly $\langle \lambda^{-1}_S, \lambda_{S'} \rangle$. To check functoriality, suppose we have $(f, g) : (S, S') \to (A, A')$ and $(f', g') : (A, A') \to (B, B')$ in $\C \times \C^\op$. Then:
  \begin{align*}
    \iota(f', g') \circ \iota(f, g)
    &= \langle \lambda^{-1}_B f', g' \lambda_{B'} \rangle \circ \langle \lambda^{-1}_A f, g \lambda_{A'} \rangle \\
    &= \langle (I\otimes (\lambda^{-1}_B f'))\lambda^{-1}_A f, g \lambda_{A'} (I\otimes (g' \lambda_{B'}))\rangle && \text{(By definition of $\circ$)}\\
    &= \langle (I \otimes \lambda^{-1}_B) (I \otimes f')\lambda^{-1}_A f, g \lambda_{A'} (I \otimes g')(I\otimes \lambda_{B'})\rangle && \text{(Functoriality of $I \otimes -$)}\\
    &= \langle (I\otimes \lambda^{-1}_B) \lambda^{-1}_B f' f, g g' \lambda_{B'} (I\otimes \lambda_{B'})\rangle && \text{(Naturality of $\lambda$)}\\
    &= \langle (\lambda^{-1}_I \otimes B) \lambda^{-1}_B f' f, g g' \lambda_{B'} (\lambda_I \otimes B')\rangle && \text{(Unitality of action)} \\
    &= \langle \lambda^{-1}_B f' f, g g' \lambda_{B'} (\lambda_I \otimes B') (\lambda^{-1}_I \otimes B') \rangle && \text{(Coend relation)}  \\
    &= \langle \lambda^{-1}_B f'f, g g' \lambda_{B'} \rangle \\
    &= \iota(f'f, gg')
  \end{align*}
  Graphically, there is not much to do:
  \begin{center}
    \input{diagrams/iota-functorial-left.tikz}
    \qquad \raisebox{0.3cm}{$=$} \qquad
    \input{diagrams/iota-functorial-right.tikz}
  \end{center}

@@

\begin{theorem}
  $\Optic_\C$ is symmetric monoidal, where $(S, S') \otimes (T, T') = (S \otimes T, S' \otimes T')$, the unit object is $(I, I)$, and the action on a pair of morphisms $\langle l, r \rangle : S \hto A$ and $\langle l', r' \rangle : T \hto B$ is given by:
  \begin{center}
    \input{diagrams/tensor-on-morphisms.tikz}
  \end{center}
\end{theorem}
\begin{proof}
  Equationally, suppose the two optics have complements $M$ and $N$ respectively. Then:
  \begin{align*}
    \langle l, r \rangle \otimes \langle l', r' \rangle &:= \langle (M \otimes s_{A,N} \otimes B)(l \otimes l'), (r \otimes r')(M \otimes s_{A',N} \otimes B') \rangle
  \end{align*}
  This is well-defined, as the calculus of string diagrams allows us to combine the coend relation on each optic to establish the coend relation for the compound diagram. \todo{draw this?}

  Let us check functoriality of $\otimes$. Suppose we have optics:
  \begin{align*}
    \langle l_1, r_1 \rangle : (S_1, S_1') &\hto (S_2, S_2') \\
    \langle l_2, r_2 \rangle : (S_2, S_2') &\hto (S_3, S_3') \\
    \langle p_1, q_1 \rangle : (T_1, T_1') &\hto (T_2, T_2') \\
    \langle p_2, q_2 \rangle : (T_2, T_2') &\hto (T_3, T_3')
  \end{align*}
  The string diagram for $(\langle l_2, r_2 \rangle \circ \langle l_1, r_1 \rangle) \otimes (\langle p_2, q_2 \rangle \circ \langle p_1, q_1 \rangle)$ is:
  \begin{center}
    \input{diagrams/tensor-functorial-left.tikz}
  \end{center}
  And for $(\langle l_2, r_2 \rangle \otimes \langle p_2, q_2 \rangle) \circ (\langle l_1, r_1 \rangle \otimes \langle p_1, q_1 \rangle)$ is:
  \begin{center}
    \input{diagrams/tensor-functorial-right.tikz}
  \end{center}
  These two diagrams are equivalent. We can use the naturality of the symmetry morphism to push $l_2$ and $r_2$ through to be next to $l_1$ and $r_1$ respectively. This creates two extra twists that can be cancelled in the center of the diagram.

  %%%% Note: The following is the above argument written equationally:
                                             %                                              For future convenience, note that
                                             %                                              \langle l, r \rangle \otimes (T, T') &= \langle (M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T') \rangle \\
      %               &= \langle (\lambda^{-1}_M \otimes A \otimes T)(M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T')(\lambda_M \otimes A' \otimes T') \rangle \\
      %               &= \langle l \otimes T, r \otimes T' \rangle
                        %                         Now we check functoriality. Suppose we have:
                        %                         \langle l_1, r_1 \rangle : (S_1, S_1') &\hto (S_2, S_2') \\
      %       \langle l_2, r_2 \rangle : (S_2, S_2') &\hto (S_3, S_3') \\
      %       \langle p_1, q_1 \rangle : (T_1, T_1') &\hto (T_2, T_2') \\
      %       \langle p_2, q_2 \rangle : (T_2, T_2') &\hto (T_3, T_3')
                                                       %                                                        with complements $M_1, M_2, N_1, N_2$ respectively. Then: \todo {(This is a mess. String diagram?)}
                                                       %                        &(\langle l_2, r_2 \rangle \otimes \langle p_2, q_2 \rangle) \circ (\langle l_1, r_1 \rangle \otimes \langle p_1, q_1 \rangle) \\
      %       = \; &\langle (M_2 \otimes s_{S_3,N_2} \otimes T_3)(l_2 \otimes p_2), (r_2 \otimes q_2)(M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
      %               &\circ \langle (M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), (r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2') \rangle \\
      %       = \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes N_1 \otimes l_2 \otimes p_2)(M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), \\
      %               &\;(r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2')(M_1 \otimes N_1 \otimes r_2 \otimes q_2)(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
      %       = \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)(M_1 \otimes  l_2 \otimes N_1 \otimes p_2)(l_1 \otimes p_1), \\
      %               &\;(r_1 \otimes q_1)(M_1 \otimes r_2 \otimes N_1 \otimes q_2)(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
      %       = \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1), \\
      %               &\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
      %       = \; & \text{\todo{(Add a $s_{M_2, N_1}$ to either side of the coend, then appeal to coherence)}} \\
      %       = \; &\langle (M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S_3} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1) ,  \\
      %               &\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S'_3} \otimes T'_3) \rangle \\
      %       = \; &\langle (M_1 \otimes l_2)l_1, r_1(M_1 \otimes r_2) \rangle \otimes \langle (N_1 \otimes p_2)p_1, q_1(N_1 \otimes q_2) \rangle \\
      %       = \; &(\langle l_2, r_2 \rangle  \circ \langle l_1, r_1 \rangle) \otimes (\langle p_2, q_2 \rangle \circ \langle p_1, q_1 \rangle)

@@

There is some further useful structure in $\Optic_\C$.
\begin{proposition}
\label{prop-costates}
  The set of costates $(S, S') \hto (I, I)$ is isomorphic to $\C(S, S')$.
\end{proposition}
\begin{proof}
  \begin{align*}
    \Optic_\C((S, S'), (I, I))
    &= \int^{M \in \C} \C(S, M \otimes I) \times \C(M \otimes I, S') \\
    &\cong \int^{M \in \C} \C(S, M) \times \C(M, S') \\
    &\cong \C(S, S')
  \end{align*}
\end{proof}

@@

In this section we establish a universal property of the $\Optic$ construction. The idea is that every optic $\langle l, r\rangle : (S, S') \hto (A, A')$ consists of a morphism $S \to M \otimes A$ and the `formal dual' of a morphism $M \otimes A' \to S'$, composed with a `formal counit' that traces out the object $M$:
\begin{center}
  \input{diagrams/generic-optic-folded.tikz}
\end{center}

@@

\todo{In this section I am using semicolon to mean composition in diagrammatic order. Should I also switch the comma in $\langle l, r \rangle$ to be a $\mid$?}

>>>>

\newcommand{\rep}[2]{{\ensuremath \langle #1 \mid #2 \rangle}}
\newcommand{\repthree}[3]{{\ensuremath \langle #1 \mid #2 \mid #3 \rangle}}
\newcommand{\repfour}[4]{{\ensuremath \langle #1 \mid #2 \mid #3 \mid #4 \rangle}}

@@

If $l : S \to M \otimes A$ and $r : M \otimes A' \to S'$, we write $\rep{l}{r} : (S, S') \hto (A, A')$ for the image of this pair in $\Optic_\C((S, S'), (A, A'))$, and say that $M$ is the complement for the representative. Optics will always be written with a crossed arrow $\hto$, to distinguish them from morphisms of $\C$.

\begin{example}
  For any three objects $M, A, A' \in \C$, there is the \emph{tautological} optic \[t_{M,A} : (M \otimes A, M \otimes A') \hto (A, A')\] which is given by $\rep{\id_{M \otimes A}}{\id_{M \otimes A}}$.
\end{example}

\begin{remark}
  Elements of $\Optic_\C((S, S'), (A, A'))$ have an appealing
  interpretation as string diagrams with a ``hole'' missing. We draw the
  pair $\rep{l}{r}$ as
  \begin{center}
    \input{diagrams/generic-optic.tikz}
  \end{center}
  reading left to right, so the left half of the diagram represents $l$ and the right half $r$. The relation imposed by the coend equates the following two diagrams:
  \begin{center}
    \input{diagrams/coend-relation-left.tikz}
    \qquad
    \input{diagrams/coend-relation-right.tikz}
  \end{center}
  We will therefore omit the vertical cut point between $l$ and $r$ in subsequent diagrams; any choice yields a representative of same optic.
\end{remark}

Two optics can be composed as follows: \todo{I give possibly too much detail here.} We wish to construct a map
\begin{align*}
  &\left(\int^{M \in \C} \C(S, M \otimes A) \times \C(M \otimes A', S')\right) \times \left(\int^{M \in \C} \C(R, M \otimes S) \times \C(M \otimes S', R')\right) \\ &
                                                                                                                                                                       \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
The product in $\Set$ preserves colimits, so in particular coends. Using this fact, and the Fubini theorem for coends~\cite[Remark 1.9]{CoendCofriend}, the domain is isomorphic to
\begin{align*}
  \int^{(M, N) \in \C \times \C} \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R').
\end{align*}
So by the universal property of coends, it suffices to construct natural maps
\begin{align*}
  & \C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') \\ &
                                                                                                              \quad \to \int^{M \in \C} \C(R, M \otimes A) \times \C(M \otimes A', R').
\end{align*}
for each $M, N \in \C$. For this we use the composites
\begin{align*}
  &\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(R, N \otimes S) \times \C(N \otimes S', R')\\
  \to \,& \C(N \otimes S, N \otimes M \otimes A) \times \C(N \otimes M \otimes A', N \otimes S') \times \C(R, N \otimes S) \times \C(N \otimes S', R') && \text{(functoriality of $N \otimes  -$)} \\
  \to \,& \C(R, N \otimes  M \otimes A) \times \C(N \otimes M \otimes A', R') && \text{(composition in $\C$)} \\
  \to \,&\int^{M \in \M} \C(R, M \otimes A) \times \C(M \otimes A', R') && \text{($\copr_{N \otimes M}$)}
\end{align*}
In other words, let $\rep{l'}{r'} : (R, R') \hto (S, S')$ and $\rep{l}{r} : (S, S') \hto (A, A')$ with $M$ the complement for $\rep{l'}{r'}$. The composite is then: $\rep{l}{r} \circ \rep{l'}{r'} = \rep{(M l)l'}{r'(Mr)}$.

@@

The identity optic $(S, S') \hto (S, S')$ is given by $\rep{\lambda^{-1}_S}{\lambda_{S'}}$, where $\lambda_S : I \otimes S \to S$ is the left unitor for $S$:
\begin{center}
  \input{diagrams/identity-optic-full.tikz}
\end{center}
This dashed line above the diagram represents the unit object. As is standard, we will elide the unit object and unitors in future diagrams unless they are necessary. The diagram for the identity becomes:
\begin{center}
  \input{diagrams/identity-optic.tikz}
\end{center}

@@

  We check associativity of composition equationally. As in the definition of composition, the Fubini theorem for coends allows us to choose representatives for all three optics simultaneously. Suppose we have representatives of three optics $\rep{l_1}{r_1} : (R, R) \hto (S, S')$, $\rep{l_2}{r_2} : (S, S') \hto (A, A')$ and $\rep{l_3}{r_3} : (A, A') \hto (B, B')$, that have complements $M$, $N$ and $P$ respectively. Then:
  \begin{align*}
    (\rep{l_3}{r_3} \circ \rep{l_2, r_2}) \circ \rep{l_1}{r_1}
    &= \rep{(N \otimes l_3)l_2}{r_2(N \otimes r_3)} \circ \rep{l_1}{r_1} \\
    &= \rep{(M \otimes ((N \otimes l_3)l_2))l_1}{r_1(M \otimes (r_2(N \otimes r_3)))} \\
    &= \rep{(M \otimes N \otimes l_3)(M \otimes l_2)l_1}{r_1(M \otimes r_2)(M \otimes N \otimes r_3)} \\
    &= \rep{l_3}{r_3} \circ (\rep{(M \otimes l_2)l_1}{r_1(M \otimes r_2)}) \\
    &= \rep{l_3}{r_3} \circ (\rep{l_2}{r_2} \circ \rep{l_1}{r_1})
  \end{align*}

  For the unit laws, suppose we have $\rep{l}{r} : (S, S') \hto (A, A')$ with representative $M$. We calculate:
  \begin{align*}
    \id_{A, A'} \circ \rep{l}{r}
    &= \rep{\lambda^{-1}_A}{\lambda_{A'}} \circ \rep{l}{r} \\
    &= \rep{(M \otimes \lambda^{-1}_A) l}{r (M\otimes  \lambda_{A'}} \\
    &= \rep{(\lambda^{-1}_M \otimes  A) l}{r (\lambda_M \otimes A'} \\
    &= \rep{l}{r (\lambda_M \otimes A') (\lambda^{-1}_M \otimes A')} \\
    &= \rep{l}{r} \\
    \rep{l}{r} \circ \id_{S, S'}
    &= \rep{l}{r} \circ \rep{\lambda^{-1}_S}{\lambda_{S'}}  \\
    &= \rep{(I \otimes l)\lambda^{-1}_S}{\lambda_{S'} (I \otimes r)} \\
    &= \rep{(\lambda^{-1}_M \otimes S)l}{r (\lambda_{M} \otimes S')} \\
    &= \rep{l}{r (\lambda_{M} \otimes S')(\lambda^{-1}_M \otimes S')} \\
    &= \rep{l}{r}
  \end{align*}
  Where in both cases we have used the naturality of $\lambda$ and coend relations to move $\lambda^{-1}_M$ to the right-hand side of the coend, where it may be cancelled.
\end{proof}

@@

\begin{proposition}
  There is a functor $\iota : \C \times \C^\op \to \Optic_\C$, which is given on objects by $\iota(S, S') = (S, S')$ and for a morphism $(f, g) : (S, S') \to (A, A')$ by $\iota(f, g) = \rep{ \lambda_A^{-1} f}{ g \lambda_{A'}}$.
\end{proposition}
\begin{proof}
  Graphically, this is:
  \begin{center}
    \input{diagrams/iota.tikz}
  \end{center}

  This preserves identities, as the identity on an object $(S, S')$ in $\Optic_\M$ is defined to be exactly $\rep{ \lambda^{-1}_S}{ \lambda_{S'}}$. To check functoriality, suppose we have $(f, g) : (S, S') \to (A, A')$ and $(f', g') : (A, A') \to (B, B')$ in $\C \times \C^\op$. Then:
  \begin{align*}
    \iota(f', g') \circ \iota(f, g)
    &= \rep{ \lambda^{-1}_B f'}{ g' \lambda_{B'}} \circ \rep{ \lambda^{-1}_A f}{ g \lambda_{A'}} \\
    &= \rep{ (I\otimes (\lambda^{-1}_B f'))\lambda^{-1}_A f}{ g \lambda_{A'} (I\otimes (g' \lambda_{B'})} && \text{(By definition of $\circ$)}\\
    &= \rep{ (I \otimes \lambda^{-1}_B) (I \otimes f')\lambda^{-1}_A f}{ g \lambda_{A'} (I \otimes g')(I\otimes \lambda_{B'}} && \text{(Functoriality of $I \otimes -$)}\\
    &= \rep{ (I\otimes \lambda^{-1}_B) \lambda^{-1}_B f' f}{ g g' \lambda_{B'} (I\otimes \lambda_{B'}} && \text{(Naturality of $\lambda$)}\\
    &= \rep{ (\lambda^{-1}_I \otimes B) \lambda^{-1}_B f' f}{ g g' \lambda_{B'} (\lambda_I \otimes B'} && \text{(Unitality of action)} \\
    &= \rep{ \lambda^{-1}_B f' f}{ g g' \lambda_{B'} (\lambda_I \otimes B') (\lambda^{-1}_I \otimes B')} && \text{(Coend relation)}  \\
    &= \rep{ \lambda^{-1}_B f'f}{ g g' \lambda_{B'}} \\
    &= \iota(f'f, gg')
  \end{align*}
  Graphically, there is not much to do:
  \begin{center}
    \input{diagrams/iota-functorial-left.tikz}
    \qquad \raisebox{0.3cm}{$=$} \qquad
    \input{diagrams/iota-functorial-right.tikz}
  \end{center}

@@

\begin{theorem}
  $\Optic_\C$ is symmetric monoidal, where $(S, S') \otimes (T, T') = (S \otimes T, S' \otimes T')$, the unit object is $(I, I)$, and the action on a pair of morphisms $\rep{ l}{ r} : S \hto A$ and $\rep{ l'}{ r'} : T \hto B$ is given by:
  \begin{center}
    \input{diagrams/tensor-on-morphisms.tikz}
  \end{center}
\end{theorem}
\begin{proof}
  Equationally, suppose the two optics have complements $M$ and $N$ respectively. Then:
  \begin{align*}
    \rep{ l}{ r} \otimes \rep{ l'}{ r'} &:= \rep{(M \otimes s_{A,N} \otimes B)(l \otimes l')}{(r \otimes r')(M \otimes s_{A',N} \otimes B')}
  \end{align*}
  This is well-defined, as the calculus of string diagrams allows us to combine the coend relation on each optic to establish the coend relation for the compound diagram. \todo{draw this?}

  Let us check functoriality of $\otimes$. Suppose we have optics:
  \begin{align*}
    \rep{ l_1}{ r_1} : (S_1, S_1') &\hto (S_2, S_2') \\
    \rep{ l_2}{ r_2} : (S_2, S_2') &\hto (S_3, S_3') \\
    \rep{ p_1}{ q_1} : (T_1, T_1') &\hto (T_2, T_2') \\
    \rep{ p_2}{ q_2} : (T_2, T_2') &\hto (T_3, T_3')
  \end{align*}
  The string diagram for $(\rep{ l_2}{ r_2} \circ \rep{ l_1}{ r_1}) \otimes (\rep{ p_2}{ q_2} \circ \rep{ p_1}{ q_1})$ is:
  \begin{center}
    \input{diagrams/tensor-functorial-left.tikz}
  \end{center}
  And for $(\rep{ l_2}{ r_2} \otimes \rep{ p_2}{ q_2}) \circ (\rep{ l_1}{ r_1} \otimes \rep{ p_1}{ q_1})$ is:
  \begin{center}
    \input{diagrams/tensor-functorial-right.tikz}
  \end{center}
  These two diagrams are equivalent. We can use the naturality of the symmetry morphism to push $l_2$ and $r_2$ through to be next to $l_1$ and $r_1$ respectively. This creates two extra twists that can be cancelled in the center of the diagram.

  %%%% Note: The following is the above argument written equationally:
  % For future convenience, note that
  % \begin{align*}
  %   \langle l, r \rangle \otimes (T, T') &= \langle (M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T') \rangle \\
  %   &= \langle (\lambda^{-1}_M \otimes A \otimes T)(M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T')(\lambda_M \otimes A' \otimes T') \rangle \\
  %   &= \langle l \otimes T, r \otimes T' \rangle
  % \end{align*}
  % Now we check functoriality. Suppose we have:
  % \begin{align*}
  %   \langle l_1, r_1 \rangle : (S_1, S_1') &\hto (S_2, S_2') \\
  %   \langle l_2, r_2 \rangle : (S_2, S_2') &\hto (S_3, S_3') \\
  %   \langle p_1, q_1 \rangle : (T_1, T_1') &\hto (T_2, T_2') \\
  %   \langle p_2, q_2 \rangle : (T_2, T_2') &\hto (T_3, T_3')
  % \end{align*}
  % with complements $M_1, M_2, N_1, N_2$ respectively. Then: \todo {(This is a mess. String diagram?)}
  % \begin{align*}
  %   &(\langle l_2, r_2 \rangle \otimes \langle p_2, q_2 \rangle) \circ (\langle l_1, r_1 \rangle \otimes \langle p_1, q_1 \rangle) \\
  %   = \; &\langle (M_2 \otimes s_{S_3,N_2} \otimes T_3)(l_2 \otimes p_2), (r_2 \otimes q_2)(M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
  %   &\circ \langle (M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), (r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2') \rangle \\
  %   = \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes N_1 \otimes l_2 \otimes p_2)(M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), \\
  %   &\;(r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2')(M_1 \otimes N_1 \otimes r_2 \otimes q_2)(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
  %   = \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)(M_1 \otimes  l_2 \otimes N_1 \otimes p_2)(l_1 \otimes p_1), \\
  %   &\;(r_1 \otimes q_1)(M_1 \otimes r_2 \otimes N_1 \otimes q_2)(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
  %   = \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1), \\
  %   &\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
  %   = \; & \text{\todo{(Add a $s_{M_2, N_1}$ to either side of the coend, then appeal to coherence)}} \\
  %   = \; &\langle (M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S_3} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1) ,  \\
  %   &\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S'_3} \otimes T'_3) \rangle \\
  %   = \; &\langle (M_1 \otimes l_2)l_1, r_1(M_1 \otimes r_2) \rangle \otimes \langle (N_1 \otimes p_2)p_1, q_1(N_1 \otimes q_2) \rangle \\
  %   = \; &(\langle l_2, r_2 \rangle  \circ \langle l_1, r_1 \rangle) \otimes (\langle p_2, q_2 \rangle \circ \langle p_1, q_1 \rangle)
  % \end{align*}

@@

There is some further useful structure in $\Optic_\C$.
\begin{proposition}\label{prop-costates}
  The set of costates $(S, S') \hto (I, I)$ is isomorphic to $\C(S, S')$.
\end{proposition}
\begin{proof}
  \begin{align*}
    \Optic_\C((S, S'), (I, I))
    &= \int^{M \in \C} \C(S, M \otimes I) \times \C(M \otimes I, S') \\
    &\cong \int^{M \in \C} \C(S, M) \times \C(M, S') \\
    &\cong \C(S, S')
  \end{align*}
\end{proof}

@@

In this section we establish a universal property of the $\Optic$ construction. The idea is that every optic $\rep{ l}{ r} : (S, S') \hto (A, A')$ consists of a morphism $S \to M \otimes A$ and the `formal dual' of a morphism $M \otimes A' \to S'$, composed with a `formal counit' that traces out the object $M$:
\begin{center}
  \input{diagrams/generic-optic-folded.tikz}
\end{center}

@@

\todo{In this section I am using semicolon to mean composition in diagrammatic order. Should I also switch the comma in $\rep{ l}{ r}$ to be a $\mid$?}
